
\subsection{Process}

The initial planning and specification phase took place during in-person weekly meetings. Real-time collaboration was achieved with a whiteboard as well as Google Docs (in-person and remotely). Everyone contributed to the high-level language design.\\\\
A Git repository on GitHub was used for version control and code storage. Individually, various text editors were used for development. The compiler was written in OCaml and compiles down to OCaml.\\\\
Tests were started early and each team member contributed to the test suite as modules were completed. A test script was used to quickly run all tests during development.

\subsection{Style Guide}

Two spaces per tab\\\\
Value\_and\_type\_names\_with\_underscores\\\\
Modules And Constructors With First Letter Uppercase\\\\
TOKENS\_ALLCAPS\\\\
Pattern matching lines and other logical groupings kept aligned

\subsection{Project Timeline}

\begin{tabular}{r l}
Nov. 24 & Finish scanner and overall design\\
Dec. 13 & Complete most of parser, translator\\
Dec. 20 & Complete presentation and report\\
Dec. 22 & Squash all bugs, final report edits\\
\end{tabular}

\subsection{Roles and Responsibilities}

Everyone contributed to writing the LRM and the final document. During  actual implementation there was a lot of cross-contribution. Here are the assigned responsibilities:
\begin{center}
\begin{tabular}{ | l | p{10cm} |}
\hline
David & \begin{itemize}
			\item primary on lexer/parser
		\end{itemize} \\ 
\hline
Hans & \begin{itemize}
			\item regression tests
			\item bug squashing
	   \end{itemize} \\
\hline
Harley & \begin{itemize}
			\item translator: arrays, array printing, pattern matching
			\item parser: pattern matching, value binding
		 \end{itemize} \\
\hline
Jonathan & \begin{itemize}
			\item language conception and team leader 
			\item primary on translator	 
		   \end{itemize}  \\
\hline
\end{tabular}
\end{center}

\subsection{Tools and Languages}
OCaml was used for the compiler and bash scripts for testing and building. Various text editors were used for writing code. Git was used for version control. Email and Google Docs were used for remote discussions and real-time collaboration.

\subsection{Project Log}
\begin{tabular}{r l}
Sep. 13 & First meeting, discussing ideas\\
Sep. 20 & Second meeting, deciding on PPL idea\\
Sep. 27 & Decided language features and scope\\
Oct. 18 & Grammar decided, skeleton of scanner implemented\\
Oct. 26 & Skeleton of parser working\\
Nov. 25 & Initial tests written in advance\\
Dec. 14 & Complete most of parser, translator\\
Dec. 16 & Translation working\\
Dec. 21 & Completed tutorial and presentation\\
Dec. 22 & Wrote last examples, finished final report edits\\
\end{tabular}
